Effectful Semantics for Substructural SSA

### 2023-11-16 pldi24/paper.tex
Tap
shrink related work

\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \]
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all
  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
  \[
    \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
    = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
  \]
  \TODO{fix broken cfg version}
\end{theorem}

@@

There is a vast literature on SSA and related compiler IRs, but for
space reasons, we are forced to restrict our discussion to the most
closely related work.

Originally, \citet{kelsey-ssa-cps} showed how to embed SSA into a CPS
representation, and \citet{appel-ssa} showed informally how SSA can be
seen as a collection of mutually-tail-recursive
procedures. \citet{chakravarty-functional-ssa-2003} made Appel's
observation formal, by showing how to translate SSA into A-normal
form~\cite{anf}, and showed how to use the translation to prove the
correctness of a constant propagation analysis. These translations
were for an untyped source and target language, and
\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
translation from SSA into the lambda calculus, which also tracks the
effects of expressions. The effect tracking is handled Koka-style,
where the effect system tracks a list of effect names. (In contrast,
we track the centrality and duplicability of terms, without reference
to the specific effect.)

@@

\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
to be ordinary three-address code programs, where the types of the
variables at each program point represent the control flows from the
assignment statements reaching that program point. However, the type
information means that every well-typed program can be placed into
SSA-form by inserting $\phi$-nodes in a completely type-directed
way. This lets them avoid having to give an operational reading of
$\phi$-nodes, since they can use the standard semantics for
three-address code.

\citet{hua-explicit-ssa-2010} give another type system for SSA
programs, as well as a direct operational semantics for it, using the
standard sequential heap semantics, as well as proving progress and
preservation for their type system.

There have also been many operational semantics for SSA arising from
compiler verification efforts. \citet{barthe-compcert-ssa-2014} give
an operational semantics of SSA as part of the CompCertSSA project,
and prove that the translation from three-address code into SSA is
semantics preserving. \citet{herklotz-gsa-2023} also formalise an
extension to SSA, "gated SSA", and give it a semantics, and give
semantics-preserving translations between GSA and ordinary SSA. Going
beyond CompCertSSA, \citet{vellum} have studied the semantics of the LLVM
IR itself.

There has been much less work on denotational semantics for SSA, or
directly on its equational theory. \citet{pop-ssa-inout-2009} give an
unusual denotational model of SSA in terms of the iteration structure
of a program, which they use to better understand the loop-closing
$\phi$-nodes found both in the gated SSA representation as well as
practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
exhibit a correspondence between an operational semantics for SSA, and
an operational semantics for the call-by-push-value calculus of
\citet{cbpv}. Despite working over an operational semantics, they use
the normal-form bisimulations of \citet{lassen-bisim} to derive an
equational theory for use in justifying optimisations. Their
operational semantics does contain not contain any effects, which
means that it cannot justify effect-dependent optimisations such as
eliminating redundant writes.

Our work is the first to validate the equational theory of SSA against
a weak memory semantics. We focused on denotational models of weak
memory, since they make it easier to establish equational properties
than the operational and axiomatic models. However, for our proof to
go through, we needed a number of features of the semantics which are
sometimes difficult to establish in the weak memory world. First,
sequencing must be associative (i.e.,
$(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
support loops, including the equation that a while-loop is equal to
its unrolling. Finally, sums need to be distributive (e.g., in the
boolean case, $(A \otimes \mb{2}) \simeq (A + A)$).

As a result, we focused on the TSO semantics of \citet{sparky}, which
does have all the properties we need. For the weaker ARM-style models
(which admit load buffering), there seem to be a number of models
which are "near misses", in that they have some, but not all, the
properties we need. The event structure model of
\citet{paviotti-modular-relaxed-dep-20} is formulated in terms of
step-indexing, so satisfies all of our needed properties except the
loop unrolling property -- step-indexing means it is a refinement
rather than an equation. The pomsets with preconditions model of
\citet{jagadeesan-pwp-20} does not exactly satisfy the associativity
of semicolon, and while the pomsets with predicate transformers model
of \citet{leaky-semicolon} does satisfy associativity, it does not yet
support loops.

@@

\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \]
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\), for all 
  \(\haslb{\Gamma}{t}{\ms{L}}\), we have
  \[
    \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
    = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
  \]
\end{theorem}

>>>>

\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \]
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}  
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  \begin{itemize} 
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
    \[
      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
    \]
    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
    \[
      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
    \]
    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  \end{itemize}
\end{theorem}

@@

For space reasons we, restrict our discussion to the most closely
related work on SSA.

Originally, \citet{kelsey-ssa-cps} showed how to embed SSA into a CPS
representation, and \citet{appel-ssa} showed informally how SSA can be
seen as a collection of mutually-tail-recursive
procedures. \citet{chakravarty-functional-ssa-2003} made Appel's
observation formal, by showing how to translate SSA into A-normal
form~\cite{anf}, and used it to prove the correctness of a constant
propagation analysis. These translations were for an untyped source
and target language, and
\citet{typed-effect-ssa-rigon-torrens-vasconcellos-20} give a typed
translation from SSA into the lambda calculus, into a type-and-effect
system. In contrast, we directly track the centrality and duplicability of
terms, without tracking specific effects.

@@

\citet{ssa-types-matsuno-ohori-06} give a type theory for what appear
to be ordinary three-address code programs. However, every well-typed
program can be placed into SSA-form by inserting $\phi$-nodes in a
fully type-directed way. This lets them model SSA without any $\phi$-nodes, lets
them use the standard
semantics for three-address code.
%
\citet{hua-explicit-ssa-2010} give another type system and
direct operational semantics for SSA, and prove type safety
for it.

Many operational semantics for SSA have arisen from compiler
verification efforts. \citet{barthe-compcert-ssa-2014} give an
operational semantics as part of the CompCertSSA project, and give a
semantics-preserving translation from three-address code into
SSA. \citet{herklotz-gsa-2023} formalise "gated SSA" and give
semantics-preserving translations between it and ordinary SSA. Going
beyond CompCertSSA, \citet{vellum} have studied the semantics of the
LLVM IR itself.

There has been much less work on denotational semantics for SSA, or
directly on its equational theory. \citet{pop-ssa-inout-2009} give an
unusual denotational model of SSA in terms of the iteration structure
of a program, which they use to better understand the loop-closing
$\phi$-nodes found both in the gated SSA representation as well as
practical compilers such as GCC.  \citet{garbuzov-structural-cfg-2018}
exhibit a correspondence between an operational semantics for SSA, and
an operational semantics for call-by-push-value~\cite{cbpv}. They use
the normal-form bisimulations of \citet{lassen-bisim} to derive an
equational theory for use in justifying optimisations. Their
operational semantics is effect-free, which means that it cannot
justify optimisations such as eliminating redundant
writes.

Our work is the first to validate the equational theory of SSA against
a weak memory semantics. We focused on denotational models of weak
memory, since they make it easier to establish equational
properties. However, we need a number of features of the semantics
which are sometimes difficult to establish in the weak memory
world. First, sequencing must be associative (i.e.,
$(t_1; t_2); t_3 \equiv t_1; (t_2; t_3)$).  Second, the semantics must
support loops, including the equation that a while-loop is equal to
its unrolling. Finally, sums need to be distributive (e.g., in the
boolean case, $(A \otimes \mb{2}) \simeq (A + A)$).

As a result, we focused on the TSO semantics of \citet{sparky}, which
validates the properties we need. For ARM-style concurrency, which
admits load buffering, we only found models which are "near misses":
they have a subset of the properties we need. The event
structure model of \citet{paviotti-modular-relaxed-dep-20} is
formulated in terms of step-indexing, so satisfies all of our needed
properties except the loop unrolling property -- step-indexing means
it is a refinement rather than an equation. The pomsets with
preconditions model of \citet{jagadeesan-pwp-20} does not exactly
satisfy the associativity of semicolon, and while the pomsets with
predicate transformers model of \citet{leaky-semicolon} does satisfy
associativity, it does not yet support loops.

@@

\begin{theorem}[Semantic Substitution] 
  Given \(\issubst{\gamma}{\Theta}{\Gamma}\),
  \begin{itemize}
    \item For all \(\hasty{\Gamma}{p}{a}{A}{q}\), 
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{p}
      ;\dnt{\hasty{\Gamma}{p}{a}{A}{q}} 
      = \dnt{\hasty{\Theta}{p}{[\gamma]a}{A}{q}}
    \]
    \item For all 
      \(\haslb{\Gamma}{t}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\), 
      \(\haslb{\Theta}{[\gamma]t}{\ms{K}}\),
    \[
      \upg{\dnt{\issubst{\gamma}{\Theta}{\Gamma}}}{}
      ; \dnt{\haslb{\Gamma}{t}{\ms{L}}}
      = \dnt{\haslb{\Theta}{[\gamma]t}{\ms{K}}} 
      ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{K}}}{\ms{K}}{\ms{L}}}
    \]
    \item For all 
      \(\lhaslb{\ms{W}}{L}{\ms{L}}\), 
      \(\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}\),
      \(\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}\),
      \[
        \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
        ; \dnt{\lhaslb{\ms{W'}}{[\gamma]L}{\ms{K}}}
        = \dnt{\lhaslb{\ms{W}}{L}{\ms{L}}}
        ; \dnt{\lbsubst{\substlbs{\gamma}{\ms{L}}}{\ms{K}}{\ms{L}}}
        + \dnt{\lbsubst{\substlbs{\gamma}{\ms{W'}}}{\ms{W}}{\ms{W'}}}
      \]
  \end{itemize}
  Similarly, given \(\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}\),
  \begin{itemize} 
    \item For all \(\haslb{\Gamma}{t}{\ms{L}}\), we have
    \[
      \dnt{\haslb{\Gamma}{t}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      = \dnt{\haslb{\Gamma}{[\mc{L}]t}{\ms{K}}}
    \]
    \item For all \(\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}\), we have
    \[
      \dnt{\lhaslb{\ms{W}_{\ms{L}}}{L}{\ms{L}}};\dnt{\lbsubst{\mc{L}}{\ms{L}}{\ms{K}}}
      + \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}}
      = \dnt{\lbsubst{\mc{L}}{\ms{W}_{\ms{L}}}{\ms{W}_{\ms{K}}}};\dnt{\lhaslb{\ms{W}_{\ms{K}}}{[\mc{L}]L}{\ms{K}}}
    \]
    where all variables in \(\ms{W}\) not in \(\ms{L}\) are fresh w.r.t \(\mc{L}\).
  \end{itemize}
\end{theorem}
